The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-06 manuscript/manuscript.Rnw
discuss cow placenta

\subsection{Cow Placenta Transcriptome}

\begin{table}
  \tiny
<<top.50.cow.expressed.genes,echo=FALSE,results="asis">>=
top.cow.genes <- top.genes.table(species.name="bos taurus")
print(xtable(top.cow.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.cow.genes,"Gene symbol")
@ 
\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
mitochondrial, or whose human ortholog is a housekeeping gene in
  \textit{Bos taurus}}
\label{tab:top50_cow_placenta}
\end{table}

The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
the dog are shown in Table 5 with a list of all genes in Supplemental
Table X. The gene with the highest expression (\ac{FPKM}=12488.8) was the
miRNA gene bta-mir-2887-1. This gene occurs in several copies in the
cow genome, and it was unclear as to how many or to which of 12
bta-mir-2887 genes this mapped (Supplemental Table X). A previous
study indicated that this miRNA gene was expressed in the bovine ovary
(Hossain et al., 2009). Several members of the ruminant-specific
pregnancy-associated glycoprotein family show especially high
expression in the cow placenta (Table 4, Supplemental Table X)
including PAG2 (\ac{FPKM}=7255.34), PAG10 (2076.42), BT.83780 (1793.72),
PAG11 (962.19), PAG12 (469.751), BT.88769 (408.584), PAG1 (301.61),
PAG4 (186.496), PAG6 (157.348), PAG8 (153.225), and PAG16 (103.573).
In addition, 14 other PAGs were expressed with \ac{FPKM}<40 (Supplemental
Table X). Other genes with high expression include genes involved with
regulation of growth (IGF2, PHLDA2), imprinted genes (GNAS, DLK1,
PHLDA2), and CSH1, a gene that is a member of a ruminant specific
expansion of placental prolactins and not directly related to the gene
of the same name in hominids. Other ruminant-specific placental
prolactins also show high expression in the cow placenta including
PRP1 (632.766), PRP5 (396.832), PRP6 (345.296), PRP2 (195.771), in
addition to 7 other placental prolactins with expression of \ac{FPKM}<51
(Table 5; Supplemental Table X). SSLP1 (also called SOLD1), a gene
expressed in the cotyledonary villi of the ruminiant placenta and
involved in cell proliferation (Ushizawa et al., 2009, 2010), also
shows evidence of high expression (\ac{FPKM}=1001.45). There is some
placental expression of the estrogen synthase gene (CYP19A1) in cows
(75.1346), and comparatively greater expression of HSD3B1 (\ac{FPKM}=
157.622).

>>>>

\subsection{\cow Placenta Transcriptome}

\begin{table}
  \tiny
<<top.50.cow.expressed.genes,echo=FALSE,results="asis">>=
top.cow.genes <- top.genes.table(species.name="bos taurus")
print(xtable(top.cow.genes),
      booktabs=TRUE,
      floating=FALSE,
      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
setkey(top.cow.genes,"Gene symbol")
@ 
\caption{\tableexplanation{\cow}{tab:all_bos_taurus_placenta}}
\label{tab:top50_cow_placenta}
\end{table}

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of the \cow are shown in \cref{tab:top50_cow_placenta} with a
list of all genes in Supplemental \cref{tab:all_bos_taurus_placenta}.
The gene with the highest expression was \textit{TMSB10} %
(\Sexpr{gene.expression("bos taurus","TMSB10")}), and
\textit{TMSB10} was also highly expressed %
(\Sexpr{gene.expression("bos taurus","TMSB4")}). As in
\dog{}, members of the annexin complex were also highly expressed
(\Sexpr{gene.expression("bos taurus","S100G",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ANXA2",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","S100A2",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","S100A11",complete=TRUE)}), as well as %
at least one of a set of four miRNA with 100\% identity
(\Sexpr{gene.expression("bos taurus","ENSBTAG00000047046",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000047154",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000045710",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000046351",complete=TRUE)}). 
%
As these are novel miRNA, there is a possibility that the assignment
of the short reads to these four miRNA is due to miss-annotation of
the \cow genome.
% 
Several members of the ruminant-specific pregnancy-associated
glycoprotein family show especially high expression in \cow placenta
(\cref{tab:top50_cow_placenta}, \cref{fig:pag2_evolution_tree}, and
Supplemental \cref{tab:all_bos_taurus_placenta}) including
%
\Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000022759",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000013214",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000019226",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","ENSBTAG00000024611",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PAG12",complete=TRUE)}. %
%
28 members of this family have $FPKM \ge 1$, and this family of
genes has greatly expanded in ruminants (\cow and \sheep{}) although
the expression levels of this family are higher in \cow than in \sheep
\cref{fig:pag2_evolution_tree}. Other genes with high expression
include genes involved with regulation of growth (\textit{IGF2},
\textit{PHLDA2}), imprinted genes (\textit{DLK1}, \textit{PHLDA2}),
and \textit{CSH1}, a gene that is a member of a ruminant specific
expansion of placental prolactins and not directly related to the gene
of the same name in hominids \ref{fig:prolactin_evolution_tree}. Other
ruminant-specific placental prolactins also show high expression in
the cow placenta including %
\Sexpr{gene.expression("bos taurus","PRP1",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP5",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP6",complete=TRUE)}, %
\Sexpr{gene.expression("bos taurus","PRP8",complete=TRUE)}, and %
\Sexpr{gene.expression("bos taurus","PRP7",complete=TRUE)}, %
with a total of 14 with $\mathrm{FPKM} \ge 1$.
%
\textit{SSLP1} (\textit{SOLD1} in \sheep{}), a gene expressed in the
cotyledonary villi of the ruminiant placenta and involved in cell
proliferation
\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}, also
shows evidence of high expression %
(\Sexpr{gene.expression("bos taurus","SSLP1")}). %
There is some placental expression of the estrogen synthase gene
(\textit{CYP19A1}) in cows %
(\Sexpr{gene.expression("bos taurus","CYP19A1")}), and
comparatively greater expression of \textit{HSD3B1} %
(\Sexpr{gene.expression("bos taurus","HSD3B1")}).

\begin{figure}
<<pag2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PAG2" & species=="bos taurus",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
                 gene.tree.table,gene.tree.expression,min.fpkm=1)
@ 
\caption{Expression of genes related to \cow \textit{PAG2} with $\mathrm{FPKM} \ge 1$}
\label{fig:pag2_evolution_tree}
\end{figure}
